Hyperplane Elimination for Quickly Enumerating Local Optima

### 2016-01-08 Hyperplane Enumeration/main.tex
Finished my edit, back down to 16 pages

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
%\institute{BEACON Center for the Study of Evolution in Action,\\
%Michigan State University, U.S.A.\\
%brianwgoldman@acm.org, punch@msu.edu}
\institute{Department\\ Organization\\ Email}

@@

\begin{abstract}
Examining the properties of local optima is a common method for understanding
combinatorial-problem landscapes.
Unfortunately,
exhaustive algorithms for finding local optima are limited to very small problem sizes.
We propose a method for exploiting problem structure
to skip hyperplanes
that cannot contain local optima, allowing runtime to scale with
the number of local optima instead of with the landscape size.
We prove optimality for linear problems and Concatenated Traps, and we provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
We further refine this method to find solutions
that cannot be improved by flipping $r$ or fewer bits, which counterintuitively
can reduce total runtime. While previous methods were limited to
landscapes with at most $2^{34}$ binary strings, hyperplane elimination can enumerate the same problems with
$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
%We believe this increase in size will help further our understanding of complex real-world landscapes.

@@

We use two common variants of NKq in our experiments which specify how the $K$ additional
variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
on variable indices $[i..(i+k) \bmod N]$. In Random NKq, $f_i$ depends on $x_i$ and a random
set of $K$ unique variables that does not include $x_i$. While Random NKq is NP-Hard,
the structure of Adjacent NKq allows for a polynomial time solution~\cite{wright:2000:solvingnk}.
Adjacent NKq is therefore often used to test a search algorithm's ability to
find the known global optimum.

@@

\section{Gray-Box Enumeration of Mk Landscapes}
\label{sec-gray-box}
When considered as a black box, the process of finding all local optima
in a landscape requires $\Omega(N2^N)$ time. This is because each
of the $2^N$ solutions must be compared with each of its $N$ neighbors.
Extending this method to look for solutions that cannot be improved by
flipping $r$ or fewer bits requires $\Omega(N^r2^N)$ time. However,
by exploiting Gray-Box optimization methods, previous work~\cite{ochoa:2015:crossovernetworks}
was able to find all $r$-bit local optima of Mk Landscapes in \BigO{2^N} for any small,
fixed value of $r$.

@@

\section{Hyperplane Elimination}
\begin{figure}
  \centering
  \includegraphics[width=.75\columnwidth,height=\textheight,keepaspectratio]{Enumerate}
  %\includegraphicsfit{Enumerate}
  \caption{Example change of ordering. The gray variables represent all dependencies
           for move $m_i$ which flips variable $F$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
  \label{fig-enumerate}
\end{figure}

Due to the limited non-linearity of the Gray-Box domain, it is possible to exclude large
parts of the search space without missing any local optima.
Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a Black-Box
domain, enumeration would progress as a binary counter, treating $index$ 0 (symbol $A$ in the solution) as
the least significant bit. This ordering ensures that before changing $index$ $i$, all possible settings of $index$
0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i-1$ positions
can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$, which flips variable $F$, is a fitness improvement
when enumeration starts (all variables set to 0). Due to the known relationships between variables,
we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
Therefore, until one of those four variables is modified, the solution cannot be a local optimum.
As a result, the hyperplane **0*00*0 cannot contain a local optima and can
therefore be eliminated from consideration during enumeration.
More generally, if at any point during enumeration
there exists a fitness-improving move, no local optima can exist until at least one
dependency of that move is modified. Any hyperplane that has all of a fitness-improving
move's dependencies fixed cannot contain any local optima.

@@

This knowledge can be exploited to skip parts of the enumeration,
as shown in Algorithm~\ref{alg-enumerate}.
Initially all variables are set to 0 and
each fitness-improving move is put into a table $move\_bin$
based on that move's lowest $index$ dependency. This is the first $index$ that
can be modified by enumeration that can potentially change the fitness effect of making that move.
In order to determine how much of the enumeration can be skipped, we must find
the highest $index$ in $move\_bin$ that contains a fitness-improving move,
as done by Line~\ref{alg-enumerate-bincheck}. If no move is fitness-improving,
then a local optimum has been found.

Algorithm~\ref{alg-enumerate} works by finding the highest $index$ in $move\_bin$
that is not empty and then adding a 1 to that $index$ in $solution$. Initially
all bins could contain a fitness-improving move, so $index$ starts at $N-1$.
If at any point all bins are empty, then the solution is added to the list of
local optima $found$. Algorithm~\ref{alg-enumerate} then
adds a 1 to $index$, using the loop on Line~\ref{alg-enumerate-counter}
to perform carry operations and Line~\ref{alg-enumerate-insert} to create the new 1 value.
Iteration stops when the carry exceeds the solution length.

@@

%TODO Reviewer 1: Explain how data structures allow this to be O(N).
We perform this reordering in a greedy fashion, such that the move with the
least-unmapped dependencies has all of its remaining dependencies mapped to the
highest remaining indices.
%New
Before iteration, each move is binned based on its number of dependencies, and mappings
are created to convert a move to its bin location and a variable to the
moves that depend on that variable. Each of these requires \BigO{N} time
as you must consider for each move (\BigO{N}) all variables (\BigO{1})
in all subfunctions (\BigO{1}) it overlaps.
The algorithm then iteratively queries the move bins in ascending order
(requiring at most $kM/N=$\BigO{1} steps) until the first non-empty bin
is found. All dependencies of that move are assigned positions, and
all moves that depend on that variable are updated in the move bin
(requiring at most \BigO{1} amortized updates). Each iteration finishes
a move, meaning after $M$ iterations the process is complete. Therefore
reordering takes \BigO{N} time.
%Old
%With proper bookkeeping, this algorithm requires
%\BigO{N} time.
We argue below that the optimal bit ordering for skipping solutions
is a greedy ordering.

@@

Consider the optimal ordering $P^\star$ that is not greedy and that has the minimum value for $\hat{i}$.
Swapping $P_{\hat{i}}^\star$ and $P_{\hat{i}+1}^\star$ cannot change the minimum dependency of any other moves.
Performing the swap causes $P_{\hat{i}}^\star$ to have the minimum dependency $P_{\hat{i}+1}^\star$ had before
the swap, while $P_{\hat{i}+1}^\star$'s minimum dependency is raised by at least
$D_{P^\star}(\hat{i}, P_{\hat{i}}^\star) - D_{P^\star}(\hat{i}, P_{\hat{i}+1}^\star)$. This value
cannot be negative, meaning that we have now constructed a solution which either skips more
solutions than $P^\star$ or has a lower $\hat{i}$ than $P^\star$, contradicting our assertions.
Therefore, there cannot exist an optimal ordering that is not greedy.
\end{proof}

@@

\begin{proof}
Consider a problem with move dependencies $m_0 = \{A, B\}$, $m_1 = \{C, D\}$, and $m_2 = \{C, D, E\}$.
There are two greedy orderings, $[m_0, m_1, m_2]$ and $[m_1, m_2, m_0]$. The former
skips $2^3+2^1$ solutions, while the latter skips $2^3+2^2$. This is because $m_1$ and
$m_2$ overlap. As a result, even though both are greedy, only the second is optimal.
\end{proof}

@@

\begin{proof}
When Algorithm~\ref{alg-enumerate} is applied to a linear
function, an index $i$ of $move\_bin$ is non-zero if and only if
$x_i$ disagrees with the global optimum.
Initially $index$ is set to $N-1$, and is decreased by at least one in every step after the first.
Each step will flip a bit which disagrees with the global optimum.
No carry operations can occur before finding the global optimum, meaning the
cost of each iteration is equal to the amount by which $index$ is decreased.
Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
the global optimum.

@@

%TODO Reviewer 2: State blocks are contiguous.
\begin{theorem}
\label{theorem-k-bound-general}
Algorithm~\ref{alg-enumerate} with reordering finds all $\prod_{i=0}^{M-1} |l_i|$ local optima
in at most $2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i|$ + \BigO{M} time.
\end{theorem}

@@

When looking for only $r$-bit local optima of $k$-bound separable problems,
Algorithm~\ref{alg-enumerate} continues to be within a constant of optimal.
For $r < k$, the additional moves reduce $c$ without increasing the cost
by more than a constant.
When $r \geq k$ there can only be one local optimum: the global optimum. Furthermore, as
only non-linearly related subsets of variables must be checked
for fitness improvements, no subsets can be added that are larger than $k$.
Therefore, even if $r=N-1$, the number of subsets only grows at \BigO{N}.
Because $c=1$ in this case, the cost of finding the single global optimum
only requires the \BigO{2^kM} time to find the first local optimum, which is
optimal.

\begin{theorem}
If $\forall_i |l_i| = c$ and $c>1$ Algorithm~\ref{alg-enumerate} with reordering
finds all $c^M$ local optima in at most $2^kc^M + $ \BigO{M} time.
\end{theorem}

@@

Concatenated Traps is a commonly used $k$-bound separable problem
with $c=2$ and $M=N/k$.  Algorithm~\ref{alg-enumerate} requires $2^k2^{N/k}$ time
to find all $2^{N/k}$ local optima in this problem. Note also that
in the degenerate case of $k=1$, a $k$-bound separable problem
is a linear function. As such, when $k=1$, $c=1$ and
\BigO{2^kM}~=~\BigO{N}.

@@

In order to flip the second variable in $s_i$, all of $s_{i-1}$ must be enumerated.
This is because only carry operations can result in $index$ being set to the second
variable in $s_i$. Each time $f_i$ contains a fitness-improving move, $index$ is set
to the first variable in $s_i$, which comes before all variables in $s_{i-1}$.
As a result, the time to solve this ordering becomes $T(i) = 2^{k-1}*T(i-1)+2^k$.
Therefore, if $c < 2^{k-1}$ this ordering performs much worse than optimal.

>>>>

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{BEACON Center for the Study of Evolution in Action,\\
Michigan State University, U.S.A.\\
brianwgoldman@acm.org, punch@msu.edu}
%\institute{Department\\ Organization\\ Email}

@@

\begin{abstract}
Examining the properties of local optima is a common method for understanding
combinatorial-problem landscapes.
Unfortunately,
exhaustive algorithms for finding local optima are limited to very small problem sizes.
We propose a method for exploiting problem structure
to skip hyperplanes
that cannot contain local optima, allowing runtime to scale with
the number of local optima instead of with the landscape size.
We prove optimality for linear functions and Concatenated Traps, and we provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
We further refine this method to find solutions
that cannot be improved by flipping $r$ or fewer bits, which counterintuitively
can reduce total runtime. While previous methods were limited to
landscapes with at most $2^{34}$ binary strings, hyperplane elimination can enumerate the same problems with
$2^{77}$ binary strings, and find all 4-bit local optima of problems with $2^{200}$ binary strings.
%We believe this increase in size will help further our understanding of complex real-world landscapes.

@@

We use two common variants of NKq in our experiments that specify how the $K$ additional
variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
on variable indices $[i..(i+k) \bmod N]$. In Random NKq, $f_i$ depends on $x_i$ and a random
set of $K$ unique variables that does not include $x_i$. While Random NKq is NP-Hard,
the structure of Adjacent NKq allows for a polynomial time solution~\cite{wright:2000:solvingnk}.

@@

\section{Gray-Box Enumeration of Mk Landscapes}
\label{sec-gray-box}
When considered as a black box, the process of finding all local optima
in a landscape requires $\Omega(N2^N)$ time. This is because each
of the $2^N$ solutions must be compared with each of its $N$ neighbors.
Extending this method to look for solutions that cannot be improved by
flipping $r$ or fewer bits requires $\Omega(N^r2^N)$ time. However,
by exploiting Gray-Box optimization methods, previous work~\cite{ochoa:2015:crossovernetworks}
was able to find all $r$-bit local optima of Mk Landscapes in \BigO{2^N} for small $r$.

@@

\section{Hyperplane Elimination}
\begin{figure}
  \centering
  \includegraphics[width=.75\columnwidth,height=\textheight,keepaspectratio]{Enumerate}
  %\includegraphicsfit{Enumerate}
  \caption{Example reordering. The gray variables are all dependencies
           for move $m_i$ which flips variable $F$.
           Reordering improves $m_i$'s lowest $index$ dependency from 2 to 4.
           }
  \label{fig-enumerate}
\end{figure}

Due to the limited non-linearity of the Gray-Box domain, it is possible to exclude large
parts of the search space without missing any local optima.
Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a Black-Box
domain, enumeration would progress as a binary counter, treating $index$ 0 (symbol $A$ in the solution) as
the least significant bit. This ordering ensures that before changing $index$ $i$, all possible settings of $index$
0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i$ positions
can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$, which flips variable $F$, is a fitness improvement
when enumeration starts (all variables set to 0). Due to the known relationships between variables,
we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
Therefore, until one of those four variables is modified, the solution cannot be a local optimum.
As a result, the hyperplane **0*00*0 cannot contain a local optima and can
therefore be eliminated from consideration during enumeration.
More generally, if at any point during enumeration
there exists a fitness-improving move, no local optima can exist until at least one
dependency of that move is modified.
%TODO Consider cutting this sentence
Any hyperplane that has all of a fitness-improving
move's dependencies fixed cannot contain any local optima.

@@

This knowledge can be exploited to skip parts of the enumeration,
as shown in Algorithm~\ref{alg-enumerate}.
Initially all variables are set to 0 and
each fitness-improving move is put into a table $move\_bin$
based on that move's lowest $index$ dependency. This is the first $index$ that
can be modified by enumeration that can potentially change the fitness effect of making that move.
Algorithm~\ref{alg-enumerate} works by finding the highest $index$ in $move\_bin$
that is not empty (Line~\ref{alg-enumerate-bincheck}) and then adding a 1 to that $index$ in $solution$. Initially
all bins could contain a fitness-improving move, so $index$ starts at $N-1$.
If at any point all bins are empty, then the solution is added to the list of
local optima $found$. Algorithm~\ref{alg-enumerate} then
adds a 1 to $index$, using the loop on Line~\ref{alg-enumerate-counter}
to perform carry operations and Line~\ref{alg-enumerate-insert} to create the new 1 value.
Iteration stops when the carry exceeds the solution length.

@@

%TODO Reviewer 1: Explain how data structures allow this to be O(N).
We perform this reordering in a greedy fashion, such that the move with the
least-unmapped dependencies has all of its remaining dependencies mapped to the
highest remaining indices.
%New
Before iteration, each move is binned based on its number of dependencies, and dictionaries
are created to convert a move to its bin location and a variable to the
moves that depend on that variable. Each of these requires \BigO{N} time
as for each move (\BigO{N}) you must consider all variables (\BigO{N})
in all subfunctions (\BigO{1}) it overlaps.
The algorithm then iteratively queries the move bins in ascending order
(requiring at most $kM/N=$\BigO{1} amortized steps) until the first non-empty bin
is found. All dependencies of that move are assigned positions, and
all moves that depend on that variable are updated in the move bin
(requiring at most \BigO{1} amortized updates). Each iteration finishes
a move, meaning after \BigO{N} iterations the process is complete.
%Old
%With proper bookkeeping, this algorithm requires
%\BigO{N} time. We argue below that the optimal bit ordering for skipping solutions
%is a greedy ordering.

@@

Consider the optimal ordering $P^\star$ that is not greedy and that has the minimum value for $\hat{i}$.
Swapping $P_{\hat{i}}^\star$ and $P_{\hat{i}+1}^\star$ cannot change the minimum dependency of any other moves.
Performing the swap causes $P_{\hat{i}}^\star$ to have the minimum dependency $P_{\hat{i}+1}^\star$ had before
the swap, while $P_{\hat{i}+1}^\star$'s minimum dependency is raised by at least
$D_{P^\star}(\hat{i}, P_{\hat{i}}^\star) - D_{P^\star}(\hat{i}, P_{\hat{i}+1}^\star)$. This value
cannot be negative, meaning that we have now constructed a solution that either skips more
solutions than $P^\star$ or has a lower $\hat{i}$ than $P^\star$, contradicting our assertions.
\end{proof}

@@

\begin{proof}
Consider a problem with move dependencies $m_0 = \{A, B\}$, $m_1 = \{C, D\}$, and $m_2 = \{C, D, E\}$.
There are two greedy orderings, $[m_0, m_1, m_2]$ and $[m_1, m_2, m_0]$. The former
skips $2^3+2^1$ solutions, while the latter skips $2^3+2^2$. This is because $m_1$ and
$m_2$ overlap. Even though both are greedy, only the second is optimal.
\end{proof}

@@

\begin{proof}
When Algorithm~\ref{alg-enumerate} is applied to a linear
function, an index $i$ of $move\_bin$ is non-zero if and only if
$x_i$ disagrees with the global optimum.
Initially $index$ is set to $N-1$, and is decreased by at least one in every step after the first.
Each step will flip a bit that disagrees with the global optimum.
No carry operations can occur before finding the global optimum, meaning the
cost of each iteration is equal to the amount by which $index$ is decreased.
Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
the global optimum.

@@

%TODO Reviewer 2: State blocks are contiguous.
\begin{theorem}
\label{theorem-k-bound-general}
Algorithm~\ref{alg-enumerate} with reordering finds all $\prod_{i=0}^{M-1} |l_i|$ local optima
in a $k$-bound problem
in at most $2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i| + $ \BigO{M} time.
\end{theorem}

@@

When looking for only $r$-bit local optima of $k$-bound separable problems,
Algorithm~\ref{alg-enumerate}'s complexity only increases by a constant.
For $r < k$, the additional moves may reduce $|l_i|$ without increasing the cost
by more than a constant.
When $r \geq k$ there can only be one local optimum: the global optimum. Furthermore, as
only non-linearly related subsets of variables must be checked
for fitness improvements, no subsets can be added that are larger than $k$.
Therefore, even if $r=N-1$, the number of subsets only grows at \BigO{N}.
Because all $|l_i|=1$ in this case, the cost of finding the single global optimum
only requires the \BigO{M} time, which is
optimal.

\begin{theorem}
If $\forall_i |l_i| = c$ and $c>1$ in a $k$-bound problem, Algorithm~\ref{alg-enumerate} with reordering
finds all $c^M$ local optima in at most $2^kc^M + $ \BigO{M} time.
\end{theorem}

@@

Concatenated Traps is a commonly used $k$-bound separable problem
with $c=2$ and $M=N/k$.  Algorithm~\ref{alg-enumerate} requires $2^k2^{N/k}$ time
to find all $2^{N/k}$ local optima in this problem. Note also that
in the degenerate case of $k=1$, a $k$-bound separable problem
is a linear function. As such, when $k=1$ and $c=1$,
$2^kc^M + $ \BigO{M}~=~\BigO{N}.

@@

In order to flip the second variable in $s_i$, all of $s_{i-1}$ must be enumerated.
This is because only carry operations can result in $index$ being set to the second
variable in $s_i$. Each time $f_i$ contains a fitness-improving move, $index$ is set
to the first variable in $s_i$, which comes before all variables in $s_{i-1}$.
As a result, the time to solve this ordering becomes $T(i) = 2^{k-1}*T(i-1)+2^k$.
Therefore, if $l_i < 2^{k-1}$ this ordering performs much worse than optimal.
